---
title: "golub"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{golub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`golub` is a landmark leukemia expression dataset distributed through
Bioconductor's `multtest` package. For this vignette, `uccdf` ships a compact
derived table called `golub_gene_panel` that keeps a small set of highly
variable genes together with the observed ALL versus AML lineage labels for
interpretation. This gives us a robust omics example in dataframe form.

# Objective

The objective is to determine whether the reduced expression panel supports a
stable sample partition and whether that partition aligns with the major
lymphoid versus myeloid lineage split. More broadly, this article shows how
`uccdf` behaves when a classical omics matrix is summarized into a sample-level
panel for exploratory clustering.

# Data preparation

```{r}
data(golub_gene_panel)
analysis_golub <- golub_gene_panel[, c(
  "sample_id", "TCL1", "TCRB_1", "CTRL_M", "CTRL_5", "IL8", "TCRB_2", "MAL", "CTRL_3"
)]
head(analysis_golub)
```

# Analysis

```{r}
fit_golub <- fit_uccdf(
  analysis_golub,
  id_column = "sample_id",
  candidate_k = 1:4,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.9,
  col_fraction = 0.9,
  seed = 333
)

fit_golub$selection
select_k(fit_golub)
```

# Results

```{r}
golub_assign <- merge(augment(fit_golub), golub_gene_panel, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(golub_assign)
```

```{r}
aggregate(
  cbind(TCL1, TCRB_1, CTRL_M, CTRL_5, IL8, TCRB_2, MAL, CTRL_3, confidence) ~ cluster,
  golub_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(golub_assign$cluster, golub_assign$lineage)
round(prop.table(table(golub_assign$cluster, golub_assign$lineage), margin = 1), 3)
```

```{r}
plot_embedding(fit_golub, color_by = "selected", main = "golub latent embedding")
plot_consensus_heatmap(fit_golub, main = "golub consensus heatmap")
```

# Discussion

The selected two-cluster solution is strong and easy to interpret. The
expression summary table usually shows coordinated differences across the chosen
genes, and the lineage table typically maps those two clusters onto the major
ALL versus AML distinction with high purity. This is what we hope to see in a
reduced omics panel: the dominant biological axis remains visible even after the
matrix has been compressed into a smaller dataframe.

This article is also useful methodologically. Omics workflows often begin with a
large matrix and then move into derived sample-level summaries. `uccdf` is aimed
at that latter stage. The consensus heatmap and null-calibrated `K` selection
show that the reduced panel still carries a stable structure that is stronger
than the column-wise null baseline.

# Interpretation

For `golub`, the clusters should be interpreted as stable expression-defined
sample groups that largely track leukemia lineage. The vignette is not meant to
replace a full supervised leukemia analysis. Its role is to demonstrate that an
omics-derived dataframe can still yield a strong, reproducible consensus
partition under the same typed workflow used for the other tabular examples.
