---
title: "ChickWeight"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ChickWeight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`ChickWeight` records repeated chick body weights across time under different
diet conditions. It is a useful biological growth dataset because the numeric
trajectory is strong, but diet still provides an important categorical context.
That combination makes it a good fit for a typed consensus workflow.

# Objective

The objective is to determine whether `uccdf` recovers stable growth regimes
from weight, diet, and a coarse time band, and to inspect whether the resulting
clusters reflect biologically interpretable stages or response patterns rather
than just arbitrary slices of the repeated-measures table.

# Data preparation

```{r}
cw_df <- as.data.frame(ChickWeight)
cw_df$sample_id <- sprintf("CW%03d", seq_len(nrow(cw_df)))
cw_df$Diet <- factor(cw_df$Diet)
cw_df$time_band <- ordered(
  cut(cw_df$Time, breaks = c(-Inf, 5, 12, Inf), labels = c("early", "mid", "late")),
  levels = c("early", "mid", "late")
)

analysis_cw <- cw_df[, c("sample_id", "weight", "Diet", "time_band")]
head(analysis_cw)
```

# Analysis

```{r}
fit_cw <- fit_uccdf(
  analysis_cw,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 707
)

fit_cw$selection
select_k(fit_cw)
```

# Results

```{r}
cw_assign <- merge(augment(fit_cw), cw_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(cw_assign)
```

```{r}
aggregate(
  cbind(weight, Time, confidence) ~ cluster,
  cw_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(cw_assign$cluster, cw_assign$Diet)
table(cw_assign$cluster, cw_assign$time_band)
round(prop.table(table(cw_assign$cluster, cw_assign$time_band), margin = 1), 3)
```

```{r}
plot_embedding(fit_cw, color_by = "selected", main = "ChickWeight latent embedding")
plot_consensus_heatmap(fit_cw, main = "ChickWeight consensus heatmap")
```

# Discussion

The selected two-cluster solution usually separates an earlier lighter growth
regime from a later heavier regime, but the result is not just a copy of the
time variable. The time-band table shows enrichment, while the diet table helps
show that diet composition still differs across the clusters. That means the
partition reflects a joint pattern in developmental timing and growth response.

This is a useful biological example because repeated-measures growth data can be
clustered in many brittle ways. A stability-first summary that returns only two
supported regimes is often more helpful than a larger segmentation that
overstates fine-grained temporal variation.

# Interpretation

For `ChickWeight`, the clusters are best interpreted as stable growth-response
regimes spanning lighter earlier observations and heavier later observations,
with diet composition contributing to how the groups are organized. The result
is descriptive rather than mechanistic, but it provides a compact and defensible
summary of the growth table.
