---
title: "PimaIndiansDiabetes2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PimaIndiansDiabetes2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`PimaIndiansDiabetes2` is a well-known clinical dataset containing laboratory
and anthropometric measurements related to diabetes risk. `uccdf` ships a
derived dataframe called `pima_biomarker_panel` that keeps the biomarker-style
measurements, age, an ordinal age band, and the observed diabetes status for
post hoc interpretation.

# Objective

The objective is to determine whether these biomarker measurements support
stable patient strata and whether the resulting groups correspond to clinically
recognizable combinations of glycemic burden, adiposity, insulin levels, and age
rather than to a single threshold on one variable.

# Data preparation

```{r}
data(pima_biomarker_panel)
analysis_pima <- pima_biomarker_panel[, c(
  "sample_id", "glucose", "pressure", "triceps", "insulin", "mass", "pedigree", "age_band"
)]
head(analysis_pima)
```

# Analysis

```{r}
fit_pima <- fit_uccdf(
  analysis_pima,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 24,
  n_null = 39,
  row_fraction = 0.9,
  col_fraction = 0.9,
  seed = 124
)

fit_pima$selection
select_k(fit_pima)
```

# Results

```{r}
pima_assign <- merge(augment(fit_pima), pima_biomarker_panel, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(pima_assign)
```

```{r}
aggregate(
  cbind(glucose, pressure, triceps, insulin, mass, pedigree, age, confidence) ~ cluster,
  pima_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(pima_assign$cluster, pima_assign$age_band)
table(pima_assign$cluster, pima_assign$diabetes)
round(prop.table(table(pima_assign$cluster, pima_assign$diabetes), margin = 1), 3)
```

```{r}
plot_embedding(fit_pima, color_by = "selected", main = "PimaIndiansDiabetes2 latent embedding")
plot_consensus_heatmap(fit_pima, main = "PimaIndiansDiabetes2 consensus heatmap")
```

# Discussion

The selected multi-cluster solution is useful because the table mixes several
biomarker axes that do not collapse to one measurement. In a typical run, one
cluster concentrates higher glucose and insulin burden, another has lower
glycemic measurements with more moderate anthropometry, and additional groups
capture intermediate or age-shifted profiles. The diabetes table is helpful
here: it usually shows enrichment rather than perfect separation, which is what
we would expect from a biomarker panel rather than a supervised classifier.

This is precisely where a typed consensus workflow helps. A single clustering
run can overstate small differences in a moderately sized clinical table, while
a stability-first summary gives a more defensible view of which patient strata
are repeatedly supported by the data.

# Interpretation

For `PimaIndiansDiabetes2`, the clusters should be interpreted as stable
biomarker profiles, not as diagnostic categories. Their practical value is that
they summarize recurring metabolic patterns that can then be compared against the
observed diabetes status, age band, and individual biomarker distributions
without turning the clustering itself into a prediction model.
