---
title: "mtcars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mtcars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`mtcars` is a classic vehicle dataset with mostly continuous specifications plus
a few discrete engineering descriptors. It is a convenient real example for
mixed-table clustering in a transport context. The table is also small enough
that unstable high-`K` segmentations are easy to invent, which makes it a good
test of whether `uccdf` stays conservative.

# Objective

The objective is to identify stable vehicle segments from performance and design
features and to see whether the resulting clusters line up with transmission,
cylinder structure, and broad efficiency-versus-power tradeoffs.

# Data preparation

```{r}
cars_df <- mtcars
cars_df$sample_id <- rownames(mtcars)
cars_df$am <- factor(cars_df$am, labels = c("automatic", "manual"))
cars_df$vs <- factor(cars_df$vs, labels = c("v-shaped", "straight"))
cars_df$cyl <- ordered(cars_df$cyl, levels = c(4, 6, 8))
cars_df$gear <- ordered(cars_df$gear, levels = c(3, 4, 5))

analysis_cars <- cars_df[, c("sample_id", "mpg", "disp", "hp", "wt", "qsec", "am", "vs", "cyl", "gear")]
head(analysis_cars)
```

# Analysis

```{r}
fit_cars <- fit_uccdf(
  analysis_cars,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 24,
  n_null = 59,
  seed = 505
)

fit_cars$selection
select_k(fit_cars)
```

# Results

```{r}
cars_assign <- merge(
  augment(fit_cars),
  cars_df,
  by.x = "row_id",
  by.y = "sample_id",
  all.x = TRUE
)
head(cars_assign)
```

```{r}
aggregate(
  cbind(mpg, disp, hp, wt, qsec, confidence) ~ cluster,
  data = cars_assign,
  FUN = function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(cars_assign$cluster, cars_assign$am)
table(cars_assign$cluster, cars_assign$cyl)
round(prop.table(table(cars_assign$cluster, cars_assign$cyl), margin = 1), 3)
```

```{r}
plot_embedding(fit_cars, color_by = "selected", main = "mtcars latent embedding")
plot_consensus_heatmap(fit_cars, main = "mtcars consensus heatmap")
```

# Discussion

The selected two-cluster solution captures a broad division in vehicle design
and performance. One cluster typically contains lighter cars with better fuel
economy, smaller displacement, and lower horsepower, while the other contains
heavier and more powerful vehicles. The `am` and `cyl` tables help confirm that
this is a mechanically meaningful split rather than a numerical artifact:
automatic transmissions and larger cylinder counts often enrich the heavier
cluster, but they do not define it perfectly on their own.

This is exactly the kind of table where a simpler stable partition is more
useful than a larger but less robust segmentation. The heatmap usually shows two
clear blocks with only a small number of boundary cars, which supports the idea
that the dominant structure is a two-segment market split.

# Interpretation

For `mtcars`, the clusters are best read as stable vehicle segments such as
lighter efficiency-oriented cars versus heavier performance-oriented cars. The
article shows that even a mostly numeric engineering table benefits from the
typed consensus workflow once a few discrete descriptors are included, because
those descriptors help anchor the interpretation of the numeric split.
