---
title: "Cars93"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cars93}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
library(MASS)
```

# Background

`Cars93` contains prices, performance, origin, drivetrain, and body type for
cars sold in the early 1990s. It is a practical transport-market table with a
rich mix of numeric and categorical descriptors. It is well suited to `uccdf`
because numerical scale alone does not fully define a market segment: body type,
origin, and drivetrain all contribute to how a car is positioned.

# Objective

The aim is to identify stable market segments from the vehicle specification
table and to test whether those segments align with recognizable product
positioning differences such as economy-oriented cars, heavier higher-power
models, and intermediate mixed-use vehicles.

# Data preparation

```{r}
cars93_df <- MASS::Cars93
cars93_df$sample_id <- sprintf("C%03d", seq_len(nrow(cars93_df)))
cars93_df$Cylinders <- ordered(as.character(cars93_df$Cylinders), levels = c("3", "4", "5", "6", "8", "rotary"))

analysis_cars93 <- cars93_df[, c(
  "sample_id", "Price", "MPG.city", "MPG.highway", "Horsepower", "Weight",
  "Type", "Origin", "DriveTrain", "Cylinders"
)]
head(analysis_cars93)
```

# Analysis

```{r}
fit_cars93 <- fit_uccdf(
  analysis_cars93,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 555
)

fit_cars93$selection
select_k(fit_cars93)
```

# Results

```{r}
cars93_assign <- merge(augment(fit_cars93), cars93_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(cars93_assign)
```

```{r}
aggregate(
  cbind(Price, MPG.city, MPG.highway, Horsepower, Weight, confidence) ~ cluster,
  cars93_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(cars93_assign$cluster, cars93_assign$Type)
table(cars93_assign$cluster, cars93_assign$Origin)
table(cars93_assign$cluster, cars93_assign$DriveTrain)
round(prop.table(table(cars93_assign$cluster, cars93_assign$Type), margin = 1), 3)
```

```{r}
plot_embedding(fit_cars93, color_by = "selected", main = "Cars93 latent embedding")
plot_consensus_heatmap(fit_cars93, main = "Cars93 consensus heatmap")
```

# Discussion

The selected three-cluster solution is well matched to the dataset because the
vehicle space is visibly multi-modal. The numeric summaries usually separate a
lighter, higher-efficiency cluster from a heavier, more powerful cluster, with a
third group occupying the middle of the market. The `Type` and `DriveTrain`
tables make this easier to read: the clusters are not arbitrary numeric slices,
but mixtures of body format, engineering style, and performance level.

This is also a good example of why a mixed workflow matters. If we clustered on
price and horsepower alone, some body-type information would be lost. Here the
consensus segmentation remains anchored in numeric structure but becomes easier
to name because categorical descriptors move with the clusters rather than being
ignored.

# Interpretation

For `Cars93`, the clusters are best interpreted as stable market segments such
as economy-oriented cars, mid-range mixed vehicles, and heavier higher-power
models. The precise marketing label is less important than the fact that the
partition is stable across resamples, visible in the heatmap, and supported
against the null baseline rather than being a one-off clustering artifact.
