---
title: "bladderEset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bladderEset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.8, fig.height = 6.8, out.width = "100%")
library(uccdf)
```

# Background

`bladderEset` is distributed through the Bioconductor `bladderbatch` experiment
package and is widely used to illustrate both biological heterogeneity and
batch effects in bladder cancer expression studies. `uccdf` ships a reduced
version called `bladder_gene_panel` that keeps a small set of highly variable
probe sets together with pathology labels and the processing batch annotation.

This dataset is valuable because the interpretation is not one-dimensional. A
cluster can reflect tumor state, pathology subtype, technical batch, or a
mixture of those effects. That makes it a more realistic omics example than a
perfectly separated textbook benchmark.

# Objective

The objective is to test whether the reduced bladder expression panel supports a
stable multi-cluster solution and to understand whether the recovered groups are
better explained by pathology status, by technical batch, or by a blend of
both. This matters because practical consensus clustering is often used in
exactly this kind of confounded setting.

# Data preparation

```{r}
if (file.exists("data/bladder_gene_panel.rda")) {
  load("data/bladder_gene_panel.rda")
} else if (file.exists("../data/bladder_gene_panel.rda")) {
  load("../data/bladder_gene_panel.rda")
} else {
  data(bladder_gene_panel, package = "uccdf")
}
analysis_bladder <- bladder_gene_panel[, c(
  "sample_id", "202917_s_at", "217022_s_at", "207935_s_at", "211430_s_at",
  "202409_at", "205916_at", "214677_x_at", "212768_s_at"
)]
head(analysis_bladder)
```

```{r}
table(bladder_gene_panel$outcome, useNA = "ifany")
table(bladder_gene_panel$batch, useNA = "ifany")
```

# Analysis

```{r}
fit_bladder <- fit_uccdf(
  analysis_bladder,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.9,
  col_fraction = 0.9,
  seed = 808
)

fit_bladder$selection
select_k(fit_bladder)
```

# Results

```{r}
bladder_assign <- merge(
  augment(fit_bladder),
  bladder_gene_panel,
  by.x = "row_id",
  by.y = "sample_id",
  all.x = TRUE
)

head(bladder_assign)
```

```{r}
aggregate(
  cbind(`202917_s_at`, `217022_s_at`, `207935_s_at`, `211430_s_at`,
        `202409_at`, `205916_at`, `214677_x_at`, `212768_s_at`, confidence) ~ cluster,
  bladder_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(bladder_assign$cluster, bladder_assign$outcome)
round(prop.table(table(bladder_assign$cluster, bladder_assign$outcome), margin = 1), 3)
```

```{r}
table(bladder_assign$cluster, bladder_assign$batch)
round(prop.table(table(bladder_assign$cluster, bladder_assign$batch), margin = 1), 3)
```

```{r}
plot_embedding(fit_bladder, color_by = "selected", main = "bladderEset latent embedding")
plot_consensus_heatmap(fit_bladder, main = "bladderEset consensus heatmap")
```

# Discussion

The bladder panel is more complicated than `ALL`, and that is precisely why it
is worth keeping in the tutorial set. The selected solution is larger than two
clusters, and the interpretation tables show that no single explanatory factor
fully accounts for every cluster. Some groups are enriched for one pathology
label, some are mixed, and some show a noticeable batch preference. That means
the consensus is capturing real structure, but the structure is not purely
biological in the simple sense of one disease label per group.

This is an important practical lesson. In real omics work, a stable cluster is
not automatically a biological subtype. The bladder example shows how `uccdf`
helps with that distinction. The null calibration tells us the multi-cluster
pattern is stronger than a column-permuted baseline, while the cross-tabulation
against `outcome` and `batch` makes it possible to inspect whether the
stability is driven by interpretable pathology, by technical effects, or by a
combination of both.

The heatmap is particularly useful here. Instead of a clean two-block layout,
we usually see several consensus blocks with different levels of separation.
That visual pattern matches the selection output: the data support multiple
groups, but the boundaries are not equally sharp across all samples.

# Interpretation

For `bladderEset`, the selected clusters should be interpreted as stable
expression-defined sample strata in a partially confounded study. Some clusters
track pathology more strongly, while others appear to absorb residual batch or
mixed heterogeneity. The right takeaway is therefore not "these are definitive
subtypes," but rather "these are the recurrent structures worth checking
against pathology and batch before making a biological claim."
