---
title: "ALL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ALL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.8, fig.height = 6.8, out.width = "100%")
library(uccdf)
```

# Background

`ALL` is a classic Bioconductor microarray experiment package for acute
lymphoblastic leukemia. The raw experiment contains expression measurements for
128 patients together with immunophenotype and molecular subtype annotations.
`uccdf` ships a compact derivative called `all_gene_panel` that preserves a
small set of highly variable probe sets together with the observed lineage and
clinical context needed for interpretation.

This is a useful omics example because the dominant biological axis is not
subtle. B-lineage and T-lineage samples differ strongly, and some molecular
subtypes sit on top of that broader split. A tutorial article should therefore
do more than state that clusters exist. It should show which expression
features carry the separation, how strongly the consensus supports the split,
and how the resulting groups line up with established leukemia annotations.

# Objective

The objective is to determine whether the reduced `ALL` panel supports a stable
consensus partition and whether the selected clusters correspond to the major
B-cell versus T-cell immunophenotype distinction. A secondary objective is to
check whether molecular subtype information remains unevenly distributed across
the recovered clusters, which would suggest that the compact panel still
contains clinically interpretable structure.

# Data preparation

```{r}
if (file.exists("data/all_gene_panel.rda")) {
  load("data/all_gene_panel.rda")
} else if (file.exists("../data/all_gene_panel.rda")) {
  load("../data/all_gene_panel.rda")
} else {
  data(all_gene_panel, package = "uccdf")
}
analysis_all <- all_gene_panel[, c(
  "sample_id", "38355_at", "36638_at", "38514_at", "41214_at",
  "36108_at", "39318_at", "38096_f_at", "38319_at"
)]
head(analysis_all)
```

```{r}
table(all_gene_panel$bt, useNA = "ifany")
table(all_gene_panel$sex, useNA = "ifany")
summary(all_gene_panel$age)
```

# Analysis

```{r}
fit_all <- fit_uccdf(
  analysis_all,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.9,
  col_fraction = 0.9,
  seed = 101
)

fit_all$selection
select_k(fit_all)
```

# Results

```{r}
all_assign <- merge(
  augment(fit_all),
  all_gene_panel,
  by.x = "row_id",
  by.y = "sample_id",
  all.x = TRUE
)

head(all_assign)
```

The selected solution should be read together with the assignment table. The
key practical outputs are the selected cluster, the confidence value, and the
original annotation columns used only for interpretation.

```{r}
aggregate(
  cbind(`38355_at`, `36638_at`, `38514_at`, `41214_at`,
        `36108_at`, `39318_at`, `38096_f_at`, `38319_at`, confidence) ~ cluster,
  all_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(all_assign$cluster, all_assign$bt)
round(prop.table(table(all_assign$cluster, all_assign$bt), margin = 1), 3)
```

```{r}
sort(table(all_assign$mol_biol), decreasing = TRUE)[1:10]
round(prop.table(table(all_assign$cluster, all_assign$mol_biol), margin = 1), 3)
```

```{r}
plot_embedding(fit_all, color_by = "selected", main = "ALL latent embedding")
plot_consensus_heatmap(fit_all, main = "ALL consensus heatmap")
```

# Discussion

This dataset behaves the way a strong omics example should. The global null is
rejected, the selected solution remains supported after null calibration, and
the consensus heatmap shows a broad block structure rather than a fragile,
sample-by-sample patchwork. In practice, the selected two-cluster solution
aligns closely with the major B-lineage versus T-lineage separation. That is
important because it shows the compact panel still carries the dominant biology
even after the original expression matrix has been reduced to a small sample
table.

The subtype table adds a second layer of interpretation. Several molecular
subtypes concentrate within one lineage-defined cluster rather than being
distributed uniformly across both groups. That is exactly the pattern we would
expect if the first consensus split captures a broad developmental axis and the
remaining within-lineage heterogeneity is real but secondary. In other words,
the compact panel is not merely separating noise; it is preserving an
interpretable hierarchy of disease structure.

The article also highlights a design point of `uccdf`. The package is not
trying to replace a full leukemia workflow based on the original
high-dimensional matrix. Instead, it shows how one can carry a biologically
grounded sample-level table into a typed consensus analysis while keeping the
output defensible with a null-calibrated `K` decision.

# Interpretation

For `ALL`, the selected clusters should be interpreted as stable
expression-defined patient groups that primarily reflect the B-cell versus
T-cell immunophenotype split. The agreement with `bt` is strong enough that the
clusters are easy to explain, while the molecular subtype table reminds us that
each cluster still contains clinically meaningful internal heterogeneity. In a
real project, this would justify drilling into subtype-enriched clusters rather
than stopping at the first partition.
