---
title: "survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
library(MASS)
```

# Background

`survey` from `MASS` contains hand measurements, pulse, and lifestyle-related
categorical fields collected from students. It is a useful example of a small
social-science table with genuinely mixed feature types. Unlike the more
engineered benchmark datasets, this table is noisy, behaviorally heterogeneous,
and only partly standardized, which makes it a realistic test of whether a
consensus workflow can stay conservative on ordinary human-subject data.

# Objective

The objective is to test whether `uccdf` can recover broad participant strata
from anthropometric and lifestyle variables after removing incomplete rows. The
intended use is exploratory: we want to see whether the data support a stable
low-dimensional grouping, not to claim latent psychological classes.

# Data preparation

```{r}
survey_df <- na.omit(MASS::survey)
survey_df$sample_id <- sprintf("SV%03d", seq_len(nrow(survey_df)))

analysis_survey <- survey_df[, c("sample_id", "Sex", "Wr.Hnd", "NW.Hnd", "W.Hnd", "Pulse", "Clap", "Exer", "Smoke")]
head(analysis_survey)
```

# Analysis

```{r}
fit_survey <- fit_uccdf(
  analysis_survey,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  min_cluster_size = 10,
  gamma_small_cluster = 3,
  seed = 444
)

fit_survey$selection
select_k(fit_survey)
```

# Results

```{r}
survey_assign <- merge(augment(fit_survey), survey_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(survey_assign)
```

```{r}
aggregate(
  cbind(Wr.Hnd, NW.Hnd, Pulse, confidence) ~ cluster,
  survey_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(survey_assign$cluster, survey_assign$Sex)
table(survey_assign$cluster, survey_assign$Clap)
table(survey_assign$cluster, survey_assign$Exer)
round(prop.table(table(survey_assign$cluster, survey_assign$Sex), margin = 1), 3)
```

```{r}
plot_embedding(fit_survey, color_by = "selected", main = "survey latent embedding")
plot_consensus_heatmap(fit_survey, main = "survey consensus heatmap")
```

# Discussion

This dataset is noisier and less engineered than the others, which makes it a
better test of whether the package can handle ordinary mixed survey data. The
selected two-cluster solution is intentionally broad. In most runs it reflects a
mix of body-size variables such as writing and non-writing hand span, together
with a smaller behavioral component visible in `Clap`, `Exer`, or `Smoke`. The
sex table often shows enrichment rather than perfect separation, which is the
right outcome for a mixed observational dataset of this kind.

The important point is that the consensus solution avoids over-fragmenting the
table. Without the small-cluster penalty, larger `K` values can carve out tiny
idiosyncratic groups driven by sparse factor combinations. Here the selected
result remains coarse enough to be interpretable while still showing a stable
structure stronger than the null baseline.

# Interpretation

For `survey`, the clusters should be read as reproducible participant strata
defined by a blend of anthropometry and reported behavior. They are not
psychological or causal latent classes. The value of the analysis is that it
shows `uccdf` behaving conservatively on a messy mixed survey table, where a
stable two-group summary is often more defensible than an aggressive
high-resolution segmentation.
