---
title: "iris"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{iris}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`iris` is the canonical morphology clustering dataset. It is mostly continuous,
but it is still useful in `uccdf` because we can add a small ordinal feature and
ask how the consensus workflow behaves on a familiar real benchmark. Because the
species labels are known, the dataset is also useful for showing what a
stability-first clustering summary does when the supervised class count and the
most reproducible unsupervised structure are not identical.

# Objective

The goal is to inspect whether `uccdf` recovers a stable species-related
structure from the morphology table and to compare the resulting clusters
against the known `Species` label, with special attention to whether the method
prefers the strongest coarse split or the full three-species partition.

# Data preparation

```{r}
iris_df <- iris
iris_df$sample_id <- sprintf("IR%03d", seq_len(nrow(iris_df)))
iris_df$petal_band <- ordered(
  cut(iris_df$Petal.Length, breaks = c(-Inf, 2.5, 5, Inf), labels = c("short", "medium", "long")),
  levels = c("short", "medium", "long")
)

analysis_iris <- iris_df[, c("sample_id", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width", "petal_band")]
head(analysis_iris)
```

# Analysis

```{r}
fit_iris <- fit_uccdf(
  analysis_iris,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 24,
  n_null = 59,
  seed = 606
)

fit_iris$selection
select_k(fit_iris)
```

# Results

```{r}
iris_assign <- merge(
  augment(fit_iris),
  iris_df[, c("sample_id", "Species", "petal_band", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")],
  by.x = "row_id",
  by.y = "sample_id",
  all.x = TRUE
)
head(iris_assign)
```

```{r}
round(prop.table(table(iris_assign$cluster, iris_assign$Species), margin = 1), 3)
```

```{r}
aggregate(
  cbind(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, confidence) ~ cluster,
  data = iris_assign,
  FUN = function(x) round(mean(x, na.rm = TRUE), 3)
)
```

```{r}
plot_embedding(fit_iris, color_by = "selected", main = "iris latent embedding")
plot_consensus_heatmap(fit_iris, main = "iris consensus heatmap")
```

# Discussion

The selected two-cluster result is informative precisely because the benchmark
has three known species. The cluster-by-species table usually shows that one
consensus group is almost entirely setosa, while the second mixes versicolor and
virginica. That means the workflow is prioritizing the strongest morphology
boundary in the data, namely the clean separation of setosa from the other two
species, rather than forcing the supervised class count into the unsupervised
summary.

This is useful in practice because many real tables do not support a uniquely
correct `K`. A stable consensus summary can legitimately prefer a coarser
partition when the finer split is weaker or less reproducible, and `iris`
demonstrates that behavior in a very transparent way.

# Interpretation

That behavior should not be treated as a failure. On `iris`, `uccdf` is telling
us that the most reproducible structure in the morphology table is a two-group
separation, roughly corresponding to setosa versus non-setosa. This is a clean
example of stability-first clustering producing a defensible reduced summary,
even when a finer biologically known label set exists.
