---
title: "attitude"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{attitude}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`attitude` contains ratings of managerial behavior and workplace climate. It is
not a biomedical dataset, but it is valuable because it represents the kind of
small organizational table where analysts are tempted to over-cluster weak
structure. Most variables are numeric ratings, and we add one ordinal summary
band to preserve the mixed-table framing.

# Objective

The objective is to determine whether the ratings support stable workplace
climate groupings and, if they do, whether those groups can be interpreted in
terms of complaints, learning opportunities, raises, and overall rating level.

# Data preparation

```{r}
att_df <- as.data.frame(attitude)
att_df$sample_id <- sprintf("AT%03d", seq_len(nrow(att_df)))
att_df$rating_band <- ordered(
  cut(att_df$rating, breaks = c(-Inf, 55, 70, Inf), labels = c("low", "mid", "high")),
  levels = c("low", "mid", "high")
)

analysis_att <- att_df[, c("sample_id", "rating", "complaints", "privileges", "learning", "raises", "rating_band")]
head(analysis_att)
```

# Analysis

```{r}
fit_att <- fit_uccdf(
  analysis_att,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 808
)

fit_att$selection
select_k(fit_att)
```

# Results

```{r}
att_assign <- merge(augment(fit_att), att_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(att_assign)
```

```{r}
aggregate(
  cbind(rating, complaints, privileges, learning, raises, confidence) ~ cluster,
  att_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(att_assign$cluster, att_assign$rating_band)
round(prop.table(table(att_assign$cluster, att_assign$rating_band), margin = 1), 3)
```

```{r}
plot_embedding(fit_att, color_by = "selected", main = "attitude latent embedding")
plot_consensus_heatmap(fit_att, main = "attitude consensus heatmap")
```

# Discussion

The selected three-cluster solution is useful because it typically separates a
more positive climate regime, a middling group, and a less favorable ratings
profile. The numeric summaries are the key evidence: differences in complaints,
learning, and raises move with the clusters rather than only the overall rating.
The ordinal rating band helps show that the solution is anchored in perceived
managerial quality but not reducible to a single thresholded variable.

This is a strong example of why stability matters. On a small ratings table it
would be easy to force a higher `K` and tell an elaborate story. The consensus
workflow instead returns only the segmentation that remains reproducible across
resamples and views.

# Interpretation

For `attitude`, the clusters should be interpreted as stable workplace-climate
profiles ranging from more favorable to less favorable managerial environments.
They are descriptive summaries of the rating table, not latent organizational
types. Their value lies in producing a compact exploratory structure that can be
explained directly from the measured variables.
