---
title: "USJudgeRatings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{USJudgeRatings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`USJudgeRatings` contains lawyer ratings of judges on several professional
dimensions such as integrity, demeanor, diligence, and case-management ability.
The table is useful because it mixes correlated judicial evaluation scores with
an ordinal band derived from one anchor variable, producing a compact but
non-trivial mixed dataset.

# Objective

The objective is to determine whether the rating table supports stable judge
profiles and whether those profiles reflect broad judicial evaluation patterns
rather than noise in any single score.

# Data preparation

```{r}
judge_df <- as.data.frame(USJudgeRatings)
judge_df$sample_id <- rownames(judge_df)
judge_df$cont_band <- ordered(
  cut(judge_df$CONT, breaks = c(-Inf, 5.5, 7, Inf), labels = c("low", "mid", "high")),
  levels = c("low", "mid", "high")
)

analysis_judge <- judge_df[, c("sample_id", "CONT", "INTG", "DMNR", "DILG", "CFMG", "cont_band")]
head(analysis_judge)
```

# Analysis

```{r}
fit_judge <- fit_uccdf(
  analysis_judge,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 909
)

fit_judge$selection
select_k(fit_judge)
```

# Results

```{r}
judge_assign <- merge(augment(fit_judge), judge_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(judge_assign)
```

```{r}
aggregate(
  cbind(CONT, INTG, DMNR, DILG, CFMG, confidence) ~ cluster,
  judge_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(judge_assign$cluster, judge_assign$cont_band)
round(prop.table(table(judge_assign$cluster, judge_assign$cont_band), margin = 1), 3)
```

```{r}
plot_embedding(fit_judge, color_by = "selected", main = "USJudgeRatings latent embedding")
plot_consensus_heatmap(fit_judge, main = "USJudgeRatings consensus heatmap")
```

# Discussion

The selected two-cluster solution typically separates judges with stronger
ratings across multiple professional dimensions from those with more moderate
scores. The important point is that the separation is multivariate: integrity,
demeanor, diligence, and case-management scores tend to move together, so the
clusters are not driven by `CONT` alone even though the ordinal band is useful
for interpretation.

This analysis is a good example of a table where the dominant structure is broad
rather than fine-grained. A stability-first method is therefore appropriate,
because an aggressive high-`K` partition would be easy to over-interpret as
judicial "types" when the data may only support a coarse favorable-versus-more-
moderate distinction.

# Interpretation

For `USJudgeRatings`, the clusters should be interpreted as stable judicial
evaluation profiles defined by jointly higher versus more moderate professional
ratings. They are not normative categories of judges. Their purpose is to give a
reproducible exploratory summary of the ratings table that can be inspected with
both numeric summaries and consensus structure plots.
