---
title: "Indometh"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Indometh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`Indometh` is a pharmacokinetic dataset tracking indomethacin concentration over
time in several subjects. It is a compact medical-style table where assay values
and subject metadata naturally mix. The repeated-measures structure makes it a
useful stress test because concentration is expected to change over time, but
subject-to-subject variation still matters.

# Objective

The aim is to see whether `uccdf` can recover stable concentration-profile
regimes using concentration, subject identity, and a coarse time band. The real
question is whether the output looks like a meaningful stratification of the
concentration curve, instead of collapsing to subject labels or to a trivial
early-versus-late split.

# Data preparation

```{r}
ind_df <- Indometh
ind_df$sample_id <- sprintf("IN%03d", seq_len(nrow(ind_df)))
ind_df$Subject <- factor(ind_df$Subject)
ind_df$time_band <- ordered(
  cut(ind_df$time, breaks = c(-Inf, 2, 4, Inf), labels = c("early", "mid", "late")),
  levels = c("early", "mid", "late")
)

analysis_ind <- ind_df[, c("sample_id", "conc", "Subject", "time_band")]
head(analysis_ind)
```

# Analysis

```{r}
fit_ind <- fit_uccdf(
  analysis_ind,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 999
)

fit_ind$selection
select_k(fit_ind)
```

# Results

```{r}
ind_assign <- merge(augment(fit_ind), ind_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(ind_assign)
```

```{r}
aggregate(
  cbind(conc, time, confidence) ~ cluster,
  ind_assign,
  function(x) round(mean(x, na.rm = TRUE), 3)
)
```

```{r}
table(ind_assign$cluster, ind_assign$Subject)
table(ind_assign$cluster, ind_assign$time_band)
round(prop.table(table(ind_assign$cluster, ind_assign$time_band), margin = 1), 3)
```

```{r}
plot_embedding(fit_ind, color_by = "selected", main = "Indometh latent embedding")
plot_consensus_heatmap(fit_ind, main = "Indometh consensus heatmap")
```

# Discussion

The selected three-cluster solution is informative because the groups usually
map onto different portions of the concentration trajectory. One cluster is
typically enriched for early high-concentration points, one captures mid-curve
measurements, and one collects lower concentrations later in the record. The
time-band table makes that progression visible, but the subject table shows that
no single subject owns a cluster outright. In other words, the partition is
about recurring pharmacokinetic regimes rather than patient identity alone.

That matters in practice. A subject-only clustering would mostly summarize who
was measured, while a pure time split would add little beyond the original
study design. The consensus result is more useful because it compresses the
table into interpretable concentration states that remain stable across
resamples.

# Interpretation

For `Indometh`, the clusters should be interpreted as descriptive concentration
regimes spanning early absorption, transition, and later elimination-like
behavior. The package is not fitting a compartment model, so the output should
not be over-read as mechanistic PK inference. Its value is that it produces a
stable exploratory stratification of the repeated-measures table that can be
used for review, faceting, or downstream diagnostics.
