---
title: "CO2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CO2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`CO2` contains uptake measurements from grass plants under different treatment
and plant-type conditions. It is a compact biology dataset where continuous
response values and experimental annotations live in the same table. The key
biological feature is that uptake is observed across concentration levels, so
the dataset mixes physiological response with experimental design factors rather
than giving a single summary number per plant.

# Objective

The goal is to ask whether `uccdf` recovers stable photosynthetic response
regimes from uptake values together with plant type, treatment, and a coarse
concentration band. In other words, we want to know whether the dominant table
structure looks like a biologically meaningful response split rather than a
mechanical recreation of the full experimental design grid.

# Data preparation

```{r}
co2_df <- CO2
co2_df$sample_id <- sprintf("CO2%03d", seq_len(nrow(co2_df)))
co2_df$conc_band <- ordered(
  cut(co2_df$conc, breaks = c(-Inf, 300, 700, Inf), labels = c("low", "mid", "high")),
  levels = c("low", "mid", "high")
)

analysis_co2 <- co2_df[, c("sample_id", "uptake", "Type", "Treatment", "conc_band")]
head(analysis_co2)
```

# Analysis

```{r}
fit_co2 <- fit_uccdf(
  analysis_co2,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 222
)

fit_co2$selection
select_k(fit_co2)
```

# Results

```{r}
co2_assign <- merge(augment(fit_co2), co2_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(co2_assign)
```

```{r}
aggregate(
  cbind(uptake, conc, confidence) ~ cluster,
  co2_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(co2_assign$cluster, co2_assign$Type)
table(co2_assign$cluster, co2_assign$Treatment)
table(co2_assign$cluster, co2_assign$conc_band)
round(prop.table(table(co2_assign$cluster, co2_assign$Type), margin = 1), 3)
```

```{r}
plot_embedding(fit_co2, color_by = "selected", main = "CO2 latent embedding")
plot_consensus_heatmap(fit_co2, main = "CO2 consensus heatmap")
```

# Discussion

The selected two-cluster solution is biologically plausible because it usually
separates a higher-uptake regime from a lower-uptake regime while only partly
aligning with `Type` and `Treatment`. The concentration-band table is especially
helpful here: high-concentration observations are enriched in the stronger
uptake cluster, but they do not define it completely. That means the consensus
solution is responding to a joint pattern in response magnitude and experimental
context, not just to one categorical field.

This is exactly where a typed consensus workflow earns its keep. A purely
numeric clustering on uptake and concentration could be dominated by the
concentration axis, while a factor-heavy approach could overstate the design
variables. The current result is a compromise: it identifies broad response
modes that remain stable across resampling.

# Interpretation

For `CO2`, the recovered groups are best interpreted as stable photosynthetic
response modes. A cautious reading is that one cluster corresponds to stronger
uptake under favorable response settings, while the other collects weaker uptake
measurements across the same experimental space. The important point is not the
exact label count; it is that the table supports a reproducible two-regime
summary that can guide downstream biological visualization and exploratory
comparison.
