---
title: "InsectSprays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{InsectSprays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`InsectSprays` records insect counts under different spray conditions. It is a
small experimental dataset with a simple mixed structure: a count-like numeric
response plus a treatment factor. Despite its size, it is a useful real-data
example because the experimental factor is known and the numeric outcome has
clear practical meaning.

# Objective

The objective is to determine whether `uccdf` recovers stable efficacy regimes
from the observed insect counts and spray labels, and to check whether the
result reflects a broad low-count versus high-count separation rather than an
over-fragmented grouping of individual sprays.

# Data preparation

```{r}
sprays_df <- InsectSprays
sprays_df$sample_id <- sprintf("IS%03d", seq_len(nrow(sprays_df)))
sprays_df$count_band <- ordered(
  cut(sprays_df$count, breaks = c(-Inf, 6, 15, Inf), labels = c("low", "mid", "high")),
  levels = c("low", "mid", "high")
)

analysis_sprays <- sprays_df[, c("sample_id", "count", "spray", "count_band")]
head(analysis_sprays)
```

# Analysis

```{r}
fit_sprays <- fit_uccdf(
  analysis_sprays,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 333
)

fit_sprays$selection
select_k(fit_sprays)
```

# Results

```{r}
sprays_assign <- merge(augment(fit_sprays), sprays_df, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(sprays_assign)
```

```{r}
aggregate(
  cbind(count, confidence) ~ cluster,
  sprays_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(sprays_assign$cluster, sprays_assign$spray)
table(sprays_assign$cluster, sprays_assign$count_band)
round(prop.table(table(sprays_assign$cluster, sprays_assign$spray), margin = 1), 3)
```

```{r}
plot_embedding(fit_sprays, color_by = "selected", main = "InsectSprays latent embedding")
plot_consensus_heatmap(fit_sprays, main = "InsectSprays consensus heatmap")
```

# Discussion

The selected two-cluster solution is easy to read against the experimental
context. One cluster is typically enriched for rows with lower counts and the
better-performing sprays, while the other collects higher-count outcomes and the
weaker treatment settings. The count-band table is useful here because it makes
clear that the split is anchored in observed efficacy rather than in arbitrary
factor coding.

This dataset is also a good calibration check for the method. Because the table
is small, a single clustering run can be unstable. The consensus workflow still
finds a coherent low-count versus high-count separation, which is exactly the
kind of conservative summary we want from a stability-oriented tool.

# Interpretation

For `InsectSprays`, the resulting partition is best interpreted as a stable
contrast between more effective and less effective spray-response regimes. The
result is simple, but it is not trivial: it demonstrates that `uccdf` can
recover an interpretable consensus split even when the dataset is small and the
signal is partly encoded by a treatment factor.
