---
title: "airquality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{airquality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8.5, fig.height = 6.2, out.width = "100%")
library(uccdf)
```

# Background

`airquality` is a classic environmental dataset containing ozone, solar
radiation, wind, temperature, and calendar information measured in New York.
It is a good first real-data analysis for `uccdf` because the variables are
easy to name but not redundant: ozone and solar radiation reflect atmospheric
conditions, wind changes pollutant dispersion, temperature tracks seasonal
warming, and month encodes broad calendar position without being a full numeric
surrogate for the weather itself.

# Objective

The analytical question is not simply whether hot days differ from cool days.
Instead, the goal is to determine whether the cleaned table contains
reproducible day-level weather regimes that jointly differ in ozone burden,
radiation, wind, and seasonal timing, and to inspect whether the resulting
clusters look like coherent environmental profiles rather than a trivial month
ordering.

# Data preparation

```{r}
aq <- na.omit(airquality)
aq$sample_id <- sprintf("AQ%03d", seq_len(nrow(aq)))
aq$Month <- ordered(month.abb[aq$Month], levels = month.abb[5:9])
aq$temp_band <- ordered(
  cut(aq$Temp, breaks = c(-Inf, 75, 85, Inf), labels = c("cool", "warm", "hot")),
  levels = c("cool", "warm", "hot")
)

analysis_aq <- aq[, c("sample_id", "Ozone", "Solar.R", "Wind", "Temp", "Month", "temp_band")]
head(analysis_aq)
```

# Analysis

```{r}
fit_aq <- fit_uccdf(
  analysis_aq,
  id_column = "sample_id",
  candidate_k = 1:5,
  n_resamples = 20,
  n_null = 39,
  row_fraction = 0.85,
  col_fraction = 0.85,
  seed = 222
)

fit_aq$selection
select_k(fit_aq)
```

# Results

```{r}
aq_assign <- merge(augment(fit_aq), aq, by.x = "row_id", by.y = "sample_id", all.x = TRUE)
head(aq_assign)
```

```{r}
aggregate(
  cbind(Ozone, Solar.R, Wind, Temp, confidence) ~ cluster,
  aq_assign,
  function(x) round(mean(x, na.rm = TRUE), 2)
)
```

```{r}
table(aq_assign$cluster, aq_assign$Month)
table(aq_assign$cluster, aq_assign$temp_band)
round(prop.table(table(aq_assign$cluster, aq_assign$Month), margin = 1), 3)
```

```{r}
plot_embedding(fit_aq, color_by = "selected", main = "airquality latent embedding")
plot_consensus_heatmap(fit_aq, main = "airquality consensus heatmap")
```

# Discussion

The three-cluster solution is useful because each group has a different
environmental signature. In a typical fit, one cluster concentrates hotter days
with elevated ozone and stronger solar radiation, another cluster is defined by
higher wind and more moderate ozone despite comparable calendar timing, and the
third cluster sits between those regimes. The month table helps show that the
solution is not just "May versus August". July and August observations are
redistributed across at least two consensus groups once wind and radiation are
considered jointly.

That detail matters for interpretation. If the method were only tracing a single
temperature gradient, the latent scatter would look like one elongated cloud and
the consensus heatmap would be close to a smooth diagonal fade. Instead the
heatmap tends to show block structure with only a modest fringe of borderline
days, which is exactly what we want from a stability-first exploratory summary.

# Interpretation

For `airquality`, the clusters should be read as reproducible daily weather
profiles with different pollution-dispersion conditions. One group corresponds
to hotter and more ozone-heavy days, one to windier and lower-burden days, and
one to intermediate mixed conditions. That is a descriptive rather than causal
statement, but it is still valuable: it turns a small mixed environmental table
into a set of defensible regimes that can be reviewed, plotted, and compared
without pretending that the data support a mechanistic latent-state model.
